A formal treatment of bidirectional typing

### 2024-01-11 tex/ESOP24/S1-intro.tex
[ tex ] Fight with the page limit

This paper is structured as follows.
We first present a concrete overview of our theory using simply typed $\lambda$-calculus in \cref{sec:key-ideas}, prior to developing a general framework for specifying bidirectional type systems in \cref{sec:defs}.
Following this, we discuss mode decoration and related properties in \cref{sec:pre-synthesis}.
The main technical contribution lies in \cref{sec:type-synthesis}, where we introduce mode-correctness and bidirectional type synthesis.
Some examples other than simply typed $\lambda$-calculus are given in \Cref{sec:example}.
We discuss further developments in \cref{sec:future} and include a demonstration in \cref{sec:formalisation} for the use of our \Agda formalisation as programs.

>>>>

This paper is structured as follows.
We first present a concrete overview of our theory using simply typed $\lambda$-calculus in \cref{sec:key-ideas}, prior to developing a general framework for specifying bidirectional type systems in \cref{sec:defs}.
Following this, we discuss mode decoration and related properties in \cref{sec:pre-synthesis}.
The main technical contribution lies in \cref{sec:type-synthesis}, where we introduce mode-correctness and bidirectional type synthesis.
Some examples other than simply typed $\lambda$-calculus are given in \Cref{sec:example} and further developments are discussed in \cref{sec:future}.
\ifarxiv
A demonstration is given in \cref{sec:formalisation} for the use of our \Agda formalisation as programs.
\fi

### 2024-01-11 tex/ESOP24/S5-synthesis.tex
[ tex ] Fight with the page limit

\begin{remark}
To make the argument above sound, it is necessary to compare types and solve a unification problem.
Hence, we assume that the set~$\Xi$ of type variables has a decidable equality, thereby ensuring that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.\footnote{%
To simplify our choice, we could simply confine $\Xi$ to any set within the family of sets $\Fin(n)$ of naturals less than~$n$, given that these sets have a decidable equality and the arity of a type construct is finite.
Indeed, in our formalisation, we adopt $\Fin(n)$ as the set of type variables in the definition of $\Type_{\Sigma}$ (see \Cref{sec:formalisation} for details).
For the sake of clarity in presentation, though, we keep using named variables and just assume that $\Xi$ has a decidable equality.}
\end{remark}
We need some auxiliary definitions for the notion of extension to state the unification problem:
\begin{defn}
By an \emph{extension}\/ $\sigma \geq \rho$ of a partial substitution $\rho$ we mean that the domain $\dom(\sigma)$ of $\sigma$ contains the domain of $\rho$ and $\sigma(x) = \rho(x)$ for every\/ $x$ in $\dom(\rho)$.
  By a \emph{minimal extension}\/ $\bar{\rho}$ of $\rho$ satisfying $P$ we mean an extension $\bar{\rho} \geq \rho$ with $P(\bar{\rho})$ such that $\sigma \geq \bar{\rho}$ whenever $\sigma \geq \rho$ and $P(\sigma)$.
\end{defn}
\begin{lemma}\label{lem:unify}
  For any\/ $A$ of\/ $\Type_{\Sigma}(\Xi)$, $B$ of\/ $\Type_{\Sigma}(\emptyset)$, and a partial substitution\/ $\rho \colon \Xi \to \Type_{\Sigma}(\emptyset)$, 
  \begin{enumerate}
    \item either there is a minimal extension\/ $\bar{\rho}$ of\/ $\rho$ such that\/ $\simsub{A}{\bar{\rho}} = B$,
    \item or there is no extension\/ $\sigma$ of\/ $\rho$ such that\/ $\simsub{A}{\sigma} = B$
  \end{enumerate}
\end{lemma}
This lemma can be inferred from the correctness of first-order unification~\citep{McBride2003,McBride2003a}, or be proved directly without unification.
We are now ready for \cref{thm:bidirectional-type-synthesis-checking}:

>>>>

\begin{remark}
To make the argument above sound, it is necessary to compare types and solve a unification problem.
Hence, we assume that the set~$\Xi$ of type variables has a decidable equality, thereby ensuring that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.\footnote{%
To simplify our choice, we may confine $\Xi$ to any set within the family of sets $\Fin(n)$ of naturals less than~$n$, given that these sets have a decidable equality and the arity of a type construct is finite.
Indeed, in our formalisation, we adopt $\Fin(n)$ as the set of type variables in the definition of $\Type_{\Sigma}$\ifarxiv (see \Cref{sec:formalisation} for details)\fi.
For the sake of clarity in presentation, though, we use named variables and just assume that $\Xi$ has a decidable equality.}
\end{remark}
We need some auxiliary definitions for the notion of extension to state the unification problem:
\begin{defn}
By an \emph{extension}\/ $\sigma \geq \rho$ of a partial substitution $\rho$ we mean that the domain $\dom(\sigma)$ of $\sigma$ contains the domain of $\rho$ and $\sigma(x) = \rho(x)$ for every\/ $x$ in $\dom(\rho)$.
  By a \emph{minimal extension}\/ $\bar{\rho}$ of $\rho$ satisfying $P$ we mean an extension $\bar{\rho} \geq \rho$ with $P(\bar{\rho})$ such that $\sigma \geq \bar{\rho}$ whenever $\sigma \geq \rho$ and $P(\sigma)$.
\end{defn}
\begin{lemma}\label{lem:unify}
  For any\/ $A$ of\/ $\Type_{\Sigma}(\Xi)$, $B$ of\/ $\Type_{\Sigma}(\emptyset)$, and a partial substitution\/ $\rho \colon \Xi \to \Type_{\Sigma}(\emptyset)$, 
  \begin{enumerate}
    \item either there is a minimal extension\/ $\bar{\rho}$ of\/ $\rho$ such that\/ $\simsub{A}{\bar{\rho}} = B$,
    \item or there is no extension\/ $\sigma$ of\/ $\rho$ such that\/ $\simsub{A}{\sigma} = B$
  \end{enumerate}
\end{lemma}
This lemma can be inferred from the correctness of first-order unification~\citep{McBride2003,McBride2003a}, or be proved directly without unification.
We are now ready for \cref{thm:bidirectional-type-synthesis-checking}:

### 2024-01-11 tex/ESOP24/S6-example.tex
[ tex ] Fight with the page limit

For illustrative purposes, consider a computational calculus~\cite{Moggi1989} with additional constructs listed in \cref{tab:computational-calculus}.
The extended language has `only' 15 constructs, far fewer than a realistic programming language would have, but there are nearly 100 possible cases to consider in bidirectional type synthesis.

On the other hand, similar to a parser generator, a type-synthesiser generator only needs a specification $(\Sigma, \Omega)$ from the user to produce a corresponding synthesiser. 
Such a specification can be derived by extending $\Sigma_{\bto}$ with additional type constructs with $\mathsf{nat}$, $\mathsf{prod}$, $\mathsf{sum}$, and $\mathsf{T}$ such that
\[
  \arity({\mathsf{nat}}) = 0\qquad
  \arity({\mathsf{T}}) = 1\qquad
  \arity({\mathsf{prod}}) = \arity({\mathsf{sum}}) = 2.
\]
Types $\mathtt{nat}$, $A \times B$, $A + B$, and $T(A)$ are given as $\tyOp_{\mathsf{nat}}$, $\tyOp_{\mathsf{prod}}(A, B)$, $\tyOp_{\mathsf{sum}}(A, B)$, and $\tyOp_{\mathsf{T}}(A)$ respectively.
The signature $\Omega_{\Lambda}^{\Leftrightarrow}$ is then extended with operations listed in \cref{tab:computational-calculus}.
Mode-correctness can be derived by \cref{lem:decidability-mode-correctness} and its type synthesiser by \cref{cor:trichotomy} with the specification~$(\Sigma, \Omega)$ directly.%
\footnote{For a demonstration in \Agda, see \cref{sec:formalisation}.}

>>>>

\ifarxiv
For illustrative purposes, consider a computational calculus~\cite{Moggi1989} with additional constructs listed in \cref{tab:computational-calculus}.
The extended language has `only' 15 constructs, far fewer than a realistic programming language would have, but there are nearly 100 possible cases to consider in bidirectional type synthesis.

On the other hand, similar to a parser generator, a type-synthesiser generator only needs a specification $(\Sigma, \Omega)$ from the user to produce a corresponding synthesiser. 
Such a specification can be derived by extending $\Sigma_{\bto}$ with additional type constructs with $\mathsf{nat}$, $\mathsf{prod}$, $\mathsf{sum}$, and $\mathsf{T}$ such that
\[
  \arity({\mathsf{nat}}) = 0\qquad
  \arity({\mathsf{T}}) = 1\qquad
  \arity({\mathsf{prod}}) = \arity({\mathsf{sum}}) = 2.
\]
Types $\mathtt{nat}$, $A \times B$, $A + B$, and $T(A)$ are given as $\tyOp_{\mathsf{nat}}$, $\tyOp_{\mathsf{prod}}(A, B)$, $\tyOp_{\mathsf{sum}}(A, B)$, and $\tyOp_{\mathsf{T}}(A)$ respectively.
The signature $\Omega_{\Lambda}^{\Leftrightarrow}$ is then extended with operations listed in \cref{tab:computational-calculus}.
Mode-correctness can be derived by \cref{lem:decidability-mode-correctness} and its type synthesiser by \cref{cor:trichotomy} with the specification~$(\Sigma, \Omega)$ directly.
For a demonstration in \Agda, see \cref{sec:formalisation}.
\else
For illustrative purposes, consider a computational calculus~\cite{Moggi1989} with naturals, sums, products, and general recursion.
The extended language has `only' 15 constructs, including pairing, projections, injections, and so on, whereas there are nearly 100 possible cases to consider in bidirectional type synthesis.
However, it has far fewer constructs than a realistic programming language would have.

On the other hand, similar to a parser generator, only one specification is needed for a type-synthesiser generator from the user to produce a corresponding synthesiser. 
Such a specification can be derived by extending $(\Sigma_{\bto}, \Omega_{\Lambda}^{\Leftrightarrow})$ accordingly for additional types and constructs with mode-correctness proved by applying \cref{lem:decidability-mode-correctness}, so its type synthesiser follows from \cref{cor:trichotomy} directly.
\fi

### 2024-01-11 tex/ESOP24/S8-future.tex
[ tex ] Fight with the page limit

\LF is naturally expressive but each extension requires a different \LF and thus a different implementation to check formal \LF proofs.
Formalising \LF and its variants is at least as hard as formalising a dependent type theory, so they are mostly implemented separately from their theory and unverified.

@@

Bidirectional type synthesis plays a crucial role in handling more complex types than simple types, and we sketch how our theory can be extended to treat a broader class of languages.
First, we need a general definition of languages, discussed above, in question (\cref{subsec:simple-types,subsec:binding-sig}).
Then, this definition can be augmented with modes (\cref{subsec:bidirectional-system}) and the definition of mode-correctness (\cref{def:mode-correctness}) can be adapted by adding the mode information into signatures.
Soundness and completeness (\cref{lem:soundness-completeness}) should still hold, as they amount to the separation and combination of mode and typing information for a given raw term (in a syntax-directed formulation).
Mode decoration (\cref{sec:mode-decoration}) involves annotating a raw term with modes and marking missing annotations, and should also work.
As for the decidability of bidirectional type synthesis, we discuss two cases involving polymorphic types and dependent types below.

@@

Extending general definitions for bidirectional typing and mode derivations from \citeauthor{Hamana2011}'s work is straightforward. 
For example, the universal type $\forall \alpha.\, A$ and type abstraction in \SystemF can be specified as operations:
\[
\mathsf{all} : * \rhd [*] * \to *
\quad\text{and}\quad
\mathsf{tabs} : [*] A \rhd \left< * \right> A^{\chk} \to \tyOp_{\mathsf{all}}(\alpha. A)^{\chk}
\]
The decidability of bidirectional type synthesis (\cref{thm:bidirectional-type-synthesis-checking}) should also carry over, as no equations are imposed on types and no guessing (for type application) is required.
Adding subtyping $A \mathrel{<:} B$ to languages can be done by replacing type equality with a subtyping relation $\mathrel{<:}$ and type equality check with subtyping check, so polymorphically typed languages with subtyping such as \SystemFsub can be specified.
The main idea of bidirectional typing does not change, so it should be possible to extend the formal theory without further assumptions too.

However, explicit type application in \SystemF and \SystemFsub is impractical but its implicit version results in a \emph{stationary rule}~\cite{Leivant1986}:
\[
  \inferrule{\Gamma \vdash t \syn \forall \alpha. A}{\Gamma \vdash t \syn A[B/\alpha]}
\]
which is not syntax-directed.
By translating the above rule to subtyping, we have the \emph{instantiation problem} which amounts to guessing $B$ in $\forall \alpha. A <: A[B/\alpha]$.
A theory that accommodates various solutions to the problem is left as future work.

@@

Extending the logical framework \Dedukti~\cite{Assaf2016}, \varcitet{Felicissimo2023}{'s recent \textsf{CompLF}} is capable of defining dependent type theories with computational rules and enables generic bidirectional type synthesis.
They propose a notion of \LF signature with modes (called \emph{moded signatures} op.\ cit.), a definition of mode-correctness (called \emph{well-typed} op.\ cit.), and an algorithm of generic bidirectional type synthesis which is decidable, sound, and complete with respect to mode-decorated terms, provided that the specified language is mode-correct, and the set of computational rules is \emph{well-behaved}, i.e.\ type-preserving, confluent, and strongly normalising.
However, the last assumption is challenging to establish even for specific languages and lacks a general understanding.

@@

There are more techniques in bidirectional typing that could be formally studied in general, with one notable example being the Pfenning recipe for bidirectionalising typing rules~\citep[Section~4]{Dunfield2021}.
There are also concepts that may be hard to fully formalise, for example `annotation character'~\citep[Section~3.4]{Dunfield2021}, which is roughly about how easy it is for the user to write annotated programs, but it would be interesting to explore to what extent such concepts can be formalised.

>>>>

The \LF approach is naturally expressive but each extension requires a different framework and a different implementation to check formal \LF proofs.
Formalising \LF and its variants is at least as hard as formalising a dependent type theory, and they are mostly implemented separately from their theory and unverified.

@@

Bidirectional type synthesis plays a crucial role in handling more complex types than simple types, and we sketch how our theory can be extended to treat a broader class of languages.
First, we need a general definition of languages in question (\cref{subsec:simple-types,subsec:binding-sig}).
Then, this definition can be augmented with modes (\cref{subsec:bidirectional-system}) and the definition of mode-correctness (\cref{def:mode-correctness}) can be adapted accordingly.
Soundness and completeness (\cref{lem:soundness-completeness}) should still hold, as they amount to the separation and combination of mode and typing information for a given raw term (in a syntax-directed formulation).
Mode decoration (\cref{sec:mode-decoration}), which annotates a raw term with modes and marks missing annotations, should also work.
As for the decidability of bidirectional type synthesis, we discuss two cases involving polymorphic types and dependent types below.

@@

Extending general definitions for bidirectional typing and mode derivations from \citeauthor{Hamana2011}'s work is straightforward. 
\ifarxiv
For example, the universal type $\forall \alpha.\, A$ and type abstraction in \SystemF can be specified as operations
\[
  \mathsf{all} : * \rhd [*] * \to *
  \quad\text{and}\quad
  \mathsf{tabs} : [*] A \rhd \left< * \right> A^{\chk} \to \tyOp_{\mathsf{all}}(\alpha. A)^{\chk}.
\]
\else
For example, the universal type $\forall \alpha.\, A$ and type abstraction in \SystemF can be specified as operations $\mathsf{all} : * \rhd [*] * \to *$ and $\mathsf{tabs} : [*] A \rhd \left< * \right> A^{\chk} \to \tyOp_{\mathsf{all}}(\alpha. A)^{\chk}$.
\fi
The decidability of bidirectional type synthesis (\cref{thm:bidirectional-type-synthesis-checking}) should also carry over, as no equations are imposed on types and no guessing (for type application) is required.
Adding subtyping $A \mathrel{<:} B$ to languages can be done by replacing type equality with a subtyping relation $\mathrel{<:}$ and type equality check with subtyping check, so polymorphically typed languages with subtyping such as \SystemFsub can be specified.
The main idea of bidirectional typing does not change, so it should be possible to extend the formal theory without further assumptions too.

\ifarxiv
However, explicit type application in \SystemF and \SystemFsub is impractical but its implicit version results in a \emph{stationary rule}~\cite{Leivant1986}:
\[
  \inferrule{\Gamma \vdash t \syn \forall \alpha. A}{\Gamma \vdash t \syn A[B/\alpha]}
\]
\else
However, explicit type application in \SystemF and \SystemFsub is impractical but its implicit version results in a \emph{stationary rule}~\cite{Leivant1986}
\fi
which is not syntax-directed.
By translating the rule to subtyping, we have the \emph{instantiation problem} which amounts to guessing $B$ in $\forall \alpha. A <: A[B/\alpha]$.
A theory that accommodates various solutions to the problem is left as future work.

@@

Extending the logical framework \Dedukti~\cite{Assaf2016}, \varcitet{Felicissimo2023}{'s recent \textsf{CompLF}} is capable of defining dependent type theories with computational rules and enables generic bidirectional type synthesis.
They propose a notion of \LF signature with modes, a definition of mode-correctness (called \emph{well-typed} op.\ cit.), and an algorithm of bidirectional type synthesis.
The algorithm is decidable, sound and complete with respect to mode-decorated terms, provided that the specified language is mode-correct, and that the set of computational rules is type-preserving, confluent, and strongly normalising.
However, the last assumption is challenging to establish even for specific languages and lacks a general understanding.

@@

There are more principles and techniques in bidirectional typing that could be formally studied in general, with one notable example being the Pfenning recipe for bidirectionalising typing rules~\citep[Section~4]{Dunfield2021}.
There are also concepts that may be hard to fully formalise, for example `annotation character'~\citep[Section~3.4]{Dunfield2021}, which is roughly about how easy it is for the user to write annotated programs, but it would be interesting to explore to what extent such concepts can be formalised.
